HMM

### 2016-01-11 paper3.Rnw
need to rescale plots, working on todo, half done

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2")
@

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
\todo{do we want to call these transition probabilities (throughout)? ML: I prefer to call it this because it is consistent with explaning HMM parts.}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
analyze the movement of Florida panthers.

@@

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figures/cat1REDplot} \caption{Model Reduction Process via Individual based Model Adjusted BIC }\label{fig:BICred_plot}
\end{figure}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figures/cat1BIC} \caption{Overall Adjusted BIC Comparisons between different HMM models}
\end{figure}

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figures/cat1avg} \caption{Average step length by time of day}
\end{figure}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figures/cat1acf} \caption{ACF }
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figures/cat1states} 
\end{figure}
\end{knitrout}

>>>>

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2")
library("depmixS4")
library("dplyr")
library("reshape2")
library("plyr")
@

@@

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in behavioural mode transition 
\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in transitioning behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transitioning
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transitioning probabilities -- in particular,
transitioning probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
analyze the movement of Florida panthers.

@@

\begin{table}[]
\centering
\caption{My caption}
\label{Cat-ID}
\begin{tabular}{ccc}
 (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
  130 & 1 & 10286    \\
  131& 2  & 1111    \\
 48& 14  & 12222    \\
 94& 15  & 9999  
\end{tabular}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{figure}
<<All mod BIC plot, warning=FALSE, echo=FALSE, message=FALSE>>=
load("cat1allmod.RData")
ggplot(allmodsumdf, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ model,ncol=4)+
    scale_y_continuous()+
    scale_size_continuous( name="Num of Parameters") +
    #    geom_point(aes(size=parameters))+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States")  +
    geom_line(aes(colour=model),size=0.5) +theme_bw() +
    ggtitle("Adjusted BIC Model Comparisons")

@
\caption{Model Reduction Process via Individual based Model Adjusted BIC }
\label{fig:BICred_plot}
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

<<Adjusted BIC comparisons, warning=FALSE, echo=FALSE, message=FALSE>>=
load("cat1sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_y_continuous()+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line(aes(colour=model),size=0.5) +theme_bw() +
    ggtitle("Adjusted BIC Model Comparisons")

@

\caption{Overall Adjusted BIC Comparisons between different HMM models}
\end{figure}
\end{knitrout}

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transitioning pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

<<avg step length by time, message=FALSE, echo=FALSE, warning=FALSE>>=
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23) + theme_bw() + 
    ggtitle("Average Step Lengths at Different Hours of the Day")

@

\caption{Average step length by time of day}
\end{figure}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
<<acf plot, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line() +theme_bw() +
    ggtitle("ACF plot of BIC Selected Models")

@
\caption{ACF }
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
<<state plots, warning=FALSE, message=FALSE, echo=FALSE>>=
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
ggplot(dat,aes(log_mean,log_sd,color=states))+
    geom_point()+geom_line() +theme_bw() +
    ggtitle("State Parameters")

@
\caption{States}
\end{figure}
\end{knitrout}
